\bf \Large A Model Evaluation Study for Treatment Planning of Laser Induced Thermal Therapy

### 2015-04-22 gpuPlanningAug13.tex
Made updates

Approximately 211,000 patients present each year in the US with brain
tumors.  Of these, 38,000 are benign primary tumors, 23,000 malignant
primary tumors, and 150,000 are metastatic, originating largely from lung,
breast, and melanoma~\cite{NCIFact13,
kalkanis2010evidence,gavrilovic2005brain,bafaloukos2004treatment,Owonikoko2014}.
The average life expectancy for patients with primary and metastatic
malignancies in the brain, from time of diagnosis until death, is
approximately 12-16 months. Five year survival is among the lowest of all
cancers. Current treatment options include conventional surgery,
stereotactic radiosurgery (SRS), or chemotherapy.  
Surgical resection may be preferred for patients presenting with a single,
solitary lesion, or lesions greater than 2.5 cm - 3.0 cm in which the size
or location of the tumor causes neurological symptoms such as seizures,
headaches, cognitive or motor deficits, that can be resolved by reducing
the volume of the disease.
SRS, such as Gamma Knife\textsuperscript{\textregistered}, is typically performed on patients with multiple
tumors, tumors under 2.5 cm in diameter, and deep seated tumors~\cite{alexander1995stereotactic}.
An additional target in brain is
epileptogenic foci for patients with medically refractory epilepsy,
where MRgLITT is being considered for the surgical armamentarium for those patients~\cite{curry2012mr,Nowell2014}.
Unfortunately, patients with malignant recurrences who have reached max radiation dose
limitations and complications with surgical resection create a group of
patients with no remaining conventional treatment options;
meanwhile patients with medically refractory epilepsy have limited interventions available.
Magnetic Resonance-guided Laser Induced Thermal Therapy (MRgLITT) presents
an alternative, minimally invasive thermal ablation technique for these
groups of patients and has been safely and successfully applied to 
each~\cite{carpentieretal08,carpentier2011laser,
carpentier2012mr,torres2013stereotactic,jethwa2012magnetic,
schwarzmaier2002basic,schulze2004laser}.

Under MR guidance, the laser applicator is carefully navigated through
critical structures and placed directly into the diseased tissue to induce
ablative heating and destroy the tissue. 
Real-time thermometry of the treatment volume during
laser heating provides a mechanism by which it is possible to deliver
these therapies in both a safe and effective 
manner~\cite{rieke2008mr, denis2005magnetic,stafford2010magnetic,
woodrum2010feasibility} as well as estimate the extent of tissue
damage~\cite{hyperthermia2003basic,mcdannold2006uterine,mcnichols2004technical}. 
However, the heating induced by the laser is not constrained exclusively
to tumor tissue and nearby tissue damage is possible. 
For these procedures to progress to standard of care, 
\textit{a priori} determination of the optimal placement of the laser catheter(s) is crucial
for achieving a more conformal delivery of therapy over the target volume
with minimal comorbidity of intervening or adjacent tissue. 
%Currently, neuro-navigation software is used to guide placement.

@@

%The computer modeling approach targets a practical clinical setting of the operating room 
%or surgical suite where portable computing hardware, efficient software implementations,
%and efficient algorithm implementations are needed to deliver responsive predictions
%of the treatment outcome.  
Two distinct modeling approaches are pursued: 
(1) A GPU implementation of an unstructured hexahedral spectral element method
for predicting the bioheat transfer is developed. 
(2) A computationally inexpensive algorithm for predicting the heat transfer
from a linear superposition of analytic point source heating functions is also
presented as a reference implementation.
{\color{red}
Combined with the data, the two modeling approaches presented provide an
environment to critically evaluate model accuracy and  selection for therapy
planning of MRgLITT. 
}
%``Essentially, all models are wrong, but some are useful'', Box~\cite{box1987empirical},
Both approaches 
build intuition in the prediction by
repetitively training the underlying physics model 
to statistically match representative datasets. 
Predictions are critically evaluated in terms of solution efficiency and
accuracy for prospective treatment planning of MRgLITT procedures. 
Leave-One-Out Cross Validation (LOOCV) is used to simulate the clinical
scenario in which $N$ datasets from previous procedures are available to
calibrate the computer model. LOOCV provides an objective framework to
critically estimate the accuracy and confidence in predicting the outcome
of the procedure for the $N+1$ patient~\cite{Stone1974,Geisser1975,Kohavi1995,Browne2000,Arlot2010}.

@@

MR thermal monitoring from MRI guided stereotactic laser induced thermal
therapy (LITT) was considered in \textit{N} = \numberofpatientsLOOCV MR thermometry datasets.
The datasets were vetted for motion artifacts, low signal-to-noise, 
and catheter induced signal voids that would
spuriously reduce the modeling accuracy.
Each patient was treated with The Visualase Thermal Therapy System
(Visualase, Inc., Houston, TX).  The Visualase\textsuperscript{\textregistered}
system includes a 15 W
980 nm diode laser, a cooling pump, and a laser applicator set.
The laser applicator set is disposable and consists of 
a 400 \si{\micro\metre} core silica fiber optic with a cylindrical diffusing tip
housed within a 1.65 \si{\milli\metre} diameter saline-cooled polycarbonate cooling 
catheter~\cite{torres2013stereotactic,mcnichols2004technical}; see
Figure~\ref{LaserApplicatorOverview}.
Applicator cooling lines and laser fiber optics are connected through a
waveguide between the control room and the bore of the MR magnet. 
An MR compatible headholder is used to secure the patient's head.
The trajectory to the targeted tumor lesion was obtained 
using the Brainlab navigation system (Brainlab, Westchester, IL USA). 
A battery-powered hand drill was used to place
a threaded plastic bone anchor within the skull.
The laser applicator is secured to the threaded plastic bone anchor.
The Visualase\textsuperscript{\textregistered} system imports images from a 3D MPRAGE sequence
to verify applicator position within the lesion. 
The depth is determined by the navigation software and is input retrospectively
within this study.

@@

MR temperature imaging was performed on a 1.5 T MRI
(GE Healthcare Technologies, Waukesha, WI)
using a 2D gradient echo sequence~\cite{staffordetal04} 
(FA = \SI{30}{\degree}, FOV = $24 \times 24$ cm\textsuperscript{2}, matrix size = $256 \times 128$, 
TR/TE=37.5/20 ms, receive-only head coil,  
5 sec. per update).
%No acceleration factor was used.
The imaging plane was chosen perpendicular to the axial direction of the applicator 
and allowed monitoring of critical structure regions.
The Visualase\textsuperscript{\textregistered} workstation communicates with the MR scanner to obtain
raw DICOM imaging data during the procedure.
The temperature dependent water proton resonance frequency shift is measured
by calculating the complex phase-difference observed during heating.
The water proton resonance frequency chemically shifts to lower
frequencies with higher temperatures (caused by rupture,
stretching, bending of hydrogen bonds)~\cite{ishihara1995paf}.
The total
temperature change, $\Delta u$, is proportional to the measured
phase change, $\delta \phi$.
\[
\Delta u = \frac{ \delta \phi } { 2 \pi \alpha \cdot \gamma B_0 \cdot \text{TE}}
\]
%$\alpha = -0.0097 \left[\frac{ppm}{^oC}\right]$ 
Here $\alpha$ is the temperature sensitivity coefficient, $\gamma$ is the
gyromagnetic ratio of water, $B_0$ is the static magnetic field strength, and
TE is the echo time.  
A baseline body temperature of \textit{u}\textsubscript{0} = \SI{37.0}{\degreeCelsius} is assumed to obtain
absolute temperature.
An Arrhenius rate process model~\cite{pearce1995rate}
was used to evaluate the thermal dose resulting from the
time-temperature history of the laser exposure.
\begin{equation}\label{ArrheniusDamage}
\text{$\Omega$}(t) = 
      \int_0^t \! Ae^{\frac{-E_A}{Ru(\tau)}} \, \mathrm{d} \tau
\end{equation}
In this Arrhenius thermal dose model, the frequency factor, $A$,
and the activation energy, $E_A$, are 
experimentally determined kinetic parameters. The values for $A$ and
$E_A$ were 3.1E98 {\color{red}s\textsuperscript{-1}} and 6.28E5 {\color{red}(J/mol)},
respectively, and have been used in previous
studies~\cite{carpentieretal08,schwarzmaier1998treatment,mcnichols2004mr}. 
$R$ is the universal gas constant.  The thermal dose was assumed to
be lethal at doses $\omega$  $\geq 1$ as seen in previous
reports~\cite{carpentieretal08,mcnichols2004mr}.

@@

Prior to treatment delivery, a low power test pulse\textemdash\textit{e.g.}, 4 W for 30 s\textemdash
is applied to verify position of the diffusion fiber optic within the catheter.  
The test pulse is
sufficient to allow thermal visualization but not cause thermal damage. 
%Criteria for inclusion in our study included...
Multiple thermal imaging datasets are available per patient;
only the therapy delivery pulses are considered in this study.
A representative laser power profile used
during the therapy is shown in Figure~\ref{LaserApplicatorOverview}. 
All DICOM header information was imported into an SQLite database 
to provide efficient queries and 
organize thermometry data for reproducible analysis and processing. 
The schema provided by the Slicer~\cite{yeniaras2013design} DICOM
module was used as template for the table structure.
The object identifier (OID)
of the SQL table was used to provide anonymous references to the data.
The SQL functions,  \verb#group by#, \verb#group_concat#, and \verb#count#,
were used in quality assurance of the data
location, number of files, etc.
Metadata needed for the analysis includes:
\begin{enumerate}
\item The applicator orientation and heating region of interest (ROI)
are manually identified within the imaging datasets for input into the
computer models discussed in Sections~\ref{BioheatComputerModel} and \ref{AnalyticBioheatModel}.
\item A text file containing the relevant laser power history for each imaging
dataset is parsed and input into the simulation. The power history provides
information on the heating and cooling time intervals during the procedure.
\end{enumerate}

@@

All computations were performed on the template hexahedral mesh
shown in Figure~\ref{LaserApplicatorOverview} (b).
For each simulation, the template was registered to the observed
laser location for each patient.
The mesh consists of disjoint regions for the applicator and tissue.
A quadrilateral mesh was extruded axially along the applicator to create
the base of the hexahedral finite element mesh. The mesh for the tissue
conforms to the surface of the application and extends sufficiently far
to ensure that the boundary does not influence the heating.
The discretization consists of $N_{dof}=$844,032 total GLL nodes. 
The degrees of freedom across the volume of the applicator were removed;
the effect of the room-temperature cooling fluid which protects the laser fiber
during heating were considered through the boundary conditions at the surface
nodes. 
{\color{red}
A mesh resolution of approximately the 1mm pixel size near the applicator was used
to ensure convergence of the discretized solution~\cite{fuentesetal11a}. 
Multiple time resolutions were also evaluated to ensure convergence of the
time stepping scheme.
}

@@

The trained model's predictive ability is
evaluated by analyzing the distribution of DSC values from the $N$ iterations of
LOOCV. 
A one-sample, one-sided Wilcoxon signed rank test examines if the trained
model prediction's median exceeds DSC $\geq$ 0.7. 
One-sample calculations were computed using a threshold DSC value, $0.7$, as 
the null hypothesis, $H_0$.
\[
   H_0: DSC  = 0.7  \qquad   H_1: DSC > 0.7
\]
{\color{red}
The value chosen is a commonly accepted value in image processing
literature~\cite{yung2010quantitative,Dice1945measures,zou2004three};  
$DSC=1$ implies complete agreement between the measure and predicted damage model,
$i.e.$ the predicted and measured damages volumes completely overlap.
}
A two-sample, paired Wilcoxon signed rank test was used to compare if the two models'
prediction medians are statistically different from one another.
All statistical tests and descriptive statistics were evaluated on
GraphPad Prism 6.01 (GraphPad Software, La Jolla, CA).

@@

Given the number of available datasets, the LOOCV analysis
provides a methodology to recapitulate the clinical scenario
for treatment planning. Prior knowledge and experience is embodied
within the thermometry data from the previous $N$ ablations.
Information from the previous ablations is extracted by calibrating
a computer model to the available data. 
In this case, we calibrated our optical parameter to the thermometry data.
{\color{red}
%We are presenting this manuscript to introduce the possibility of a new
%paradigm of treatment planning via a train-and-predict scheme.  This method
%is becoming possible because of a growing number of retrospective patient
%datasets. 
%% Relatively simple models were chosen for this demonstration
%% to expedite the model training, \textit{i.e.}, optimization. 
Modeling goals are to ultimately utilize the calibrated model in optimizing
the thermal dose delivery.
The success
or failure of this paradigm is related to several factors.
First, there must be a sufficient quantity of retrospective datasets
for the training to converge. Second, the cohort of retrospective
datasets used in training must have sufficient similarity within the group
and to the prediction scenarios. Third, the model must be able 
to describe clinically relevant ablations. 
%In this investigation, none of these three factors are satisfied. 
%Nonetheless,
%we propose that model training may expedite clinically useful,
%patient-specific prediction by reducing the model complexity necessary
%and/or the need for patient-specific or population-based parameter
%measurements.

@@

The two models presented provides a canonical model selection comparison of the
trade-off between the time investment in the algorithm, efficiency of the
numerical implementation, and accuracy required for predicting the final endpoint
of the application. The two-sided, two-sample paired Wilcoxon signed rank test
indicates the predictive results' medians are significantly different and
only the transient model's median DSC performance was significantly
$\geq$ 0.7. 
{\color{red}
As a quantitative reference,
the forward solve mapping between model parameters, Table~\ref{modeldata}, 
and temperature field is considered the fundamental computational
operation of this study.
Total runtimes in the analysis are  proportional to the total number of foward solve
iterations in the optimation and LOOCV analysis. 
The run times for the forward solves in this study 
averaged 187.e3s and 11.2s for the transient model GPU implementation 
and steady state CPU implementation; respectively. 
These numbers are intended to provide intuition for the observed
practical run times on a local CPU 
% (Intel Xeon, 6 core, 2.4 GHz)  
(Intel Xeon, 6 core, 2.4 GHz, double precision peak 57 Gflops )  
workstation with attach GPU accelerators
%(NVIDIA Tesla M2070)
(NVIDIA Tesla M2070 double precision peak 515 Gflops) 
available to this study.
}
The finite element discretization of the governing equations requires significant
expertise of the GPU computing architecture as well as detailed algorithmic
understanding of both the finite element technology and matrix-vector multiply
within the iterative linear system solver to maximize floating point operation
throughput and minimizes memory transactions latencies.
As opposed to explicitly storing and reading matrix entries
from global memory, the matrix-free method recalculates the local matrix entries
needed within the linear system solve.
This memory access design pattern has demonstrated a 4-10x speedup  over
the matrix explicit methods~\cite{Muller2013,Knepley_2_2013,Medina2014}.
Meanwhile, the steady state superposition analysis, Equation~\eqn{GreenSuperPosition},
provides treatment predictions with fewer floating point operations and would be selected under an Occam
razor~\cite{jaynes2003probability} philosophy highly preferential to simplicity.
While all kernels
for the present spectral element methods and preconditioned conjugate gradient
method were hand coded in this manuscript, library implementations of the
matrix-free iterative solver approach are also
appearing~\cite{Knepley2013,Knepley2014}.

@@

The finite element discretization of the governing equations, however, 
provides significant opportunity for further
physics-based improvements including 
higher order model spherical harmonic expansions in the laser fluence 
model~\cite{carp2004radiative}. 
As seen in Figure~\ref{QuantitativeComparison}, the models used
introduce a bias in the model parameter
recovery that differs from published literature values.
{\color{red}The bias may arise from inaccuracies in the modeling assumptions, perhaps
most significant being the use of optical tissue properties that are
invariant in space and time. The literature has clear examples
where temperature/damage dependent and spatially/temporally dependent
parameters are critical to the
prediction~\cite{Mohammed05,Schutt2008}. The use of thermometry data during
only the test pulse is also expected to influence the recovered optical
properties.
The choice to use a single constant parameter for full time history of each
MRTI dataset is motivated by
the practical model training focus of this investigation.
Higher order physics
models of the fluence are be expected to provide the highest accuracy in
recovering the optical parameters during the calibration process and for
characterizing the tissue properties.

>>>>

Approximately 211,000 patients present each year in the US with brain
tumors.  Of these, 38,000 are benign primary tumors, 23,000 malignant
primary tumors, and 150,000 are metastatic, originating largely from lung,
breast, and melanoma~\cite{NCIFact13,
kalkanis2010evidence,gavrilovic2005brain,bafaloukos2004treatment,Owonikoko2014}.
The average life expectancy for patients with primary and metastatic
malignancies in the brain, from time of diagnosis until death, is
approximately 12-16 months. Five year survival is among the lowest of all
cancers. Current treatment options include conventional surgery,
stereotactic radiosurgery (SRS), or chemotherapy.  
Surgical resection may be preferred for patients presenting with a single,
solitary lesion, or lesions greater than 2.5 cm - 3.0 cm in which the size
or location of the tumor causes neurological symptoms such as seizures,
headaches, cognitive or motor deficits, that can be resolved by reducing
the volume of the disease.
SRS, such as Gamma Knife\textsuperscript{\textregistered}, is typically performed on patients with multiple
tumors, tumors under 2.5 cm in diameter, and deep seated tumors~\cite{alexander1995stereotactic}.
An additional target in brain is
epileptogenic foci for patients with medically refractory epilepsy,
where MRgLITT is being considered for the surgical armamentarium for those patients~\cite{curry2012mr,Nowell2014}.
Unfortunately, patients with malignant recurrences who have reached max{\color{red}imum} radiation dose
limitations and complications with surgical resection create a group of
patients with no remaining conventional treatment options;
meanwhile patients with medically refractory epilepsy have limited interventions available.
Magnetic resonance-guided laser induced thermal therapy (MRgLITT) presents
an alternative, minimally invasive thermal ablation technique for these
groups of patients and has been safely and successfully applied to 
each~\cite{carpentieretal08,carpentier2011laser,
carpentier2012mr,torres2013stereotactic,jethwa2012magnetic,
schwarzmaier2002basic,schulze2004laser}.

Under MR guidance, the laser applicator is carefully navigated through
critical structures and placed directly into the diseased tissue to induce
ablative heating and destroy the tissue. 
Real-time thermometry of the treatment volume during
laser heating provides a mechanism by which it is possible to deliver
these therapies in both a safe and effective 
manner~\cite{rieke2008mr, denis2005magnetic,stafford2010magnetic,
woodrum2010feasibility} as well as estimate the extent of tissue
damage~\cite{hyperthermia2003basic,mcdannold2006uterine,mcnichols2004technical}. 
However, the heating induced by the laser is not constrained exclusively
to tumor tissue and nearby tissue damage is possible. 
For these procedures to progress to standard of care, 
\textit{a priori} determination of the optimal placement of the laser catheter(s) is crucial
for achieving a more conformal delivery of therapy over the target volume
with minimal comorbidity of intervening or adjacent tissue.
{\color{red}Surgical workflows in which the operating room and MRI suite are separate~\cite{Curry2012,Tovar-Spinoza2013}
exaggerate the advantages of a predictive model.}
%Currently, neuro-navigation software is used to guide placement.

@@

%The computer modeling approach targets a practical clinical setting of the operating room 
%or surgical suite where portable computing hardware, efficient software implementations,
%and efficient algorithm implementations are needed to deliver responsive predictions
%of the treatment outcome.  
Two distinct modeling approaches are pursued: 
(1) A GPU implementation of an unstructured hexahedral spectral element method
for predicting the bioheat transfer is developed. 
(2) A computationally inexpensive algorithm for predicting the heat transfer
from a linear superposition of analytic point source heating functions is also
presented as a reference implementation.
{\color{red}
Combined with the data, the two modeling approaches presented provide an
environment to critically evaluate model accuracy and  selection for therapy
planning of MRgLITT. 
}
%``Essentially, all models are wrong, but some are useful'', Box~\cite{box1987empirical},
Both approaches 
build intuition in the prediction by
repetitively training the underlying physics model 
to statistically match representative datasets. 
Predictions are critically evaluated in terms of solution efficiency and
accuracy for prospective treatment planning of MRgLITT procedures. 
Leave-one-out cross validation (LOOCV) is used to simulate the clinical
scenario in which $N$ datasets from previous procedures are available to
calibrate the computer model. LOOCV provides an objective framework to
critically estimate the accuracy and confidence in predicting the outcome
of the procedure for the $N+1$ patient~\cite{Stone1974,Geisser1975,Kohavi1995,Browne2000,Arlot2010}.

@@

MR thermal monitoring from MRI guided stereotactic laser induced thermal
therapy (LITT) was considered in \textit{N} = \numberofpatientsLOOCV MR thermometry datasets.
The datasets were vetted for motion artifacts, low signal-to-noise, 
and catheter induced signal voids that
spuriously reduce the modeling accuracy.
Each patient was treated with The Visualase Thermal Therapy System
(Visualase, Inc., Houston, TX).  The Visualase\textsuperscript{\textregistered}
system includes a 15 W
980 nm diode laser, a cooling pump, and a laser applicator set.
The laser applicator set is disposable and consists of 
a 400 \si{\micro\metre} core silica fiber optic with a cylindrical diffusing tip
housed within a 1.65 \si{\milli\metre} diameter saline-cooled polycarbonate cooling 
catheter~\cite{torres2013stereotactic,mcnichols2004technical}; see
Figure~\ref{LaserApplicatorOverview}.
Applicator cooling lines and laser fiber optics are connected through a
waveguide between the control room and the bore of the MR magnet. 
An MR compatible headholder is used to secure the patient's head.
The trajectory to the targeted tumor lesion was obtained 
using the Brainlab navigation system (Brainlab, Westchester, IL USA). 
A battery-powered hand drill was used to place
a threaded plastic bone anchor within the skull.
The laser applicator is secured to the threaded plastic bone anchor.
The Visualase\textsuperscript{\textregistered} system imports images from a 3D MPRAGE sequence
to verify applicator position within the lesion. 
The depth is determined by the navigation software and is input retrospectively
within this study.

@@

MR temperature imaging was performed on a 1.5 T MRI
(GE Healthcare Technologies, Waukesha, WI)
using a 2D gradient echo sequence~\cite{staffordetal04} 
(FA = \SI{30}{\degree}, FOV = $24 \times 24$ cm\textsuperscript{2}, matrix size = $256 \times 128$, 
TR/TE=37.5/20 ms, receive-only head coil,  
5 sec{\color{red}onds} per update).
%No acceleration factor was used.
The imaging plane was chosen perpendicular to the axial direction of the applicator 
and allowed monitoring of critical structure regions.
The Visualase\textsuperscript{\textregistered} workstation communicates with the MR scanner to obtain
raw DICOM imaging data during the procedure.
The temperature dependent water proton resonance frequency shift is measured
by calculating the complex phase-difference observed during heating.
The water proton resonance frequency chemically shifts to lower
frequencies with higher temperatures (caused by rupture,
stretching, bending of hydrogen bonds)~\cite{ishihara1995paf}.
The total
temperature change, $\Delta u$, is proportional to the measured
phase change, $\delta \phi$.
\[
\Delta u = \frac{ \delta \phi } { 2 \pi \alpha \cdot \gamma B_0 \cdot \text{TE}}
\]
%$\alpha = -0.0097 \left[\frac{ppm}{^oC}\right]$ 
Here $\alpha$ is the temperature sensitivity coefficient, $\gamma$ is the
gyromagnetic ratio of water, $B_0$ is the static magnetic field strength, and
TE is the echo time.  
A baseline body temperature of \textit{u}\textsubscript{0} = \SI{37.0}{\degreeCelsius} is assumed to obtain
absolute temperature.
An Arrhenius rate process model~\cite{pearce1995rate}
was used to evaluate the thermal dose resulting from the
time-temperature history of the laser exposure.
\begin{equation}\label{ArrheniusDamage}
\text{$\Omega$}(t) = 
      \int_0^t \! Ae^{\frac{-E_A}{Ru(\tau)}} \, \mathrm{d} \tau
\end{equation}
In this Arrhenius thermal dose model, the frequency factor, $A$,
and the activation energy, $E_A$, are 
experimentally determined kinetic parameters. The values for $A$ and
$E_A$ were 3.1E98 {\color{red}s\textsuperscript{-1}} and 6.28E5 {\color{red}(J/mol)},
respectively, and have been used in previous
studies~\cite{carpentieretal08,schwarzmaier1998treatment,mcnichols2004mr}. 
$R$ is the universal gas constant.  The thermal dose was assumed to
be lethal at doses $\omega$  $\geq 1$ as seen in previous
reports~\cite{carpentieretal08,mcnichols2004mr}.

@@

Prior to treatment delivery, a low power test pulse\textemdash\textit{e.g.}, 4 W for 30 s\textemdash
is applied to verify position of the diffusion fiber optic within the catheter.  
The test pulse is
sufficient to allow thermal visualization but not cause thermal damage. 
%Criteria for inclusion in our study included...
Multiple thermal imaging datasets are available per patient;
only the {\color{red}therapeutic} pulses are considered in this study.
A representative laser power profile used
during the therapy is shown in Figure~\ref{LaserApplicatorOverview}. 
All DICOM header information was imported into an SQLite database 
to provide efficient queries and 
organize thermometry data for reproducible analysis and processing. 
The schema provided by the Slicer~\cite{yeniaras2013design} DICOM
module was used as template for the table structure.
The object identifier (OID)
of the SQL table was used to provide anonymous references to the data.
The SQL functions,  \verb#group by#, \verb#group_concat#, and \verb#count#,
were used in quality assurance of the data
location, number of files, etc.
Metadata needed for the analysis includes:
\begin{enumerate}
\item The applicator orientation and heating region of interest (ROI)
are manually identified within the imaging datasets for input into the
computer models discussed in Sections~\ref{BioheatComputerModel} and \ref{AnalyticBioheatModel}.
\item A text file containing the relevant laser power history for each imaging
dataset is parsed and input into the simulation. The power history provides
information on the heating and cooling time intervals during the procedure.
\end{enumerate}

@@

All computations were performed on the template hexahedral mesh
shown in Figure~\ref{LaserApplicatorOverview} (b).
For each simulation, the template was registered to the observed
laser location for each patient.
The mesh consists of disjoint regions for the applicator and tissue.
A quadrilateral mesh was extruded axially along the applicator to create
the base of the hexahedral finite element mesh. The mesh for the tissue
conforms to the surface of the application and extends sufficiently far
to ensure that the boundary does not influence the heating.
The discretization consists of $N_{dof}=$844,032 total GLL nodes. 
The degrees of freedom across the volume of the applicator were removed;
the effect of the room-temperature cooling fluid which protects the laser fiber
during heating were considered through the boundary conditions at the surface
nodes. 
{\color{red}
A mesh resolution approximating the 1 mm pixel size near the applicator was used
to ensure convergence of the discretized solution~\cite{fuentesetal11a}. 
Multiple time resolutions were also evaluated to ensure convergence of the
time stepping scheme.
}

@@

The trained model's predictive ability is
evaluated by analyzing the distribution of DSC values from the $N$ iterations of
LOOCV. 
A one-sample, one-sided Wilcoxon signed rank test examines if the trained
model prediction's median exceeds DSC $\geq$ 0.7. 
One-sample calculations were computed using a threshold DSC value, $0.7$, as 
the null hypothesis, $H_0$.
\[
   H_0: DSC  = 0.7  \qquad   H_1: DSC > 0.7
\]
{\color{red}
The value chosen is a commonly accepted value in image processing
literature~\cite{yung2010quantitative,Dice1945measures,zou2004three};  
$DSC=1$ implies complete agreement between the measure and predicted damage model,
$i.e.$, the predicted and measured damages volumes completely overlap.
}
A two-sample, paired Wilcoxon signed rank test was used to compare if the two models'
prediction medians are statistically different from one another.
All statistical tests and descriptive statistics were evaluated on
GraphPad Prism 6.01 (GraphPad Software, La Jolla, CA).

@@

Given the number of available datasets, the LOOCV analysis
provides a methodology to recapitulate the clinical scenario
for treatment planning. Prior knowledge and experience is embodied
within the thermometry data from the previous $N$ ablations.
Information from the previous ablations is extracted by calibrating
a computer model to the available data. 
In this case, we calibrate our optical parameter to the thermometry data.
{\color{red}
%We are presenting this manuscript to introduce the possibility of a new
%paradigm of treatment planning via a train-and-predict scheme.  This method
%is becoming possible because of a growing number of retrospective patient
%datasets. 
%% Relatively simple models were chosen for this demonstration
%% to expedite the model training, \textit{i.e.}, optimization. 
Modeling goals are to ultimately utilize the calibrated model in optimizing
the thermal dose delivery.
The success
or failure of this paradigm is related to several factors.
First, there must be a sufficient quantity of retrospective datasets
for the training to converge. Second, the cohort of retrospective
datasets used in training must have sufficient similarity within the group
and to the prediction scenarios. Third, the model must be able 
to describe clinically relevant ablations. 
%In this investigation, none of these three factors are satisfied. 
%Nonetheless,
%we propose that model training may expedite clinically useful,
%patient-specific prediction by reducing the model complexity necessary
%and/or the need for patient-specific or population-based parameter
%measurements.

@@

The two models presented provides a canonical model selection comparison of the
trade-off between the time investment in the algorithm, efficiency of the
numerical implementation, and accuracy required for predicting the final endpoint
of the application. The two-sided, two-sample paired Wilcoxon signed rank test
indicates the predictive results' medians are significantly different and
only the transient model's median DSC performance was significantly
$\geq$ 0.7. 
{\color{red}
As a quantitative reference,
the forward solve mapping between model parameters, Table~\ref{modeldata}, 
and temperature field is considered the fundamental computational
operation of this study.
Total runtimes in the analysis are  proportional to the total number of foward solve
iterations in the optimation and LOOCV analysis. 
The run times for the forward solves in this study 
averaged 187E3 s and 11.2 s for the transient model GPU implementation 
and steady state CPU implementation; respectively. 
These numbers are intended to provide intuition for the observed
practical run times on a local CPU 
% (Intel Xeon, 6 core, 2.4 GHz)  
(Intel Xeon, 6 core, 2.4 GHz, double precision peak 57 Gflops)  
workstation with attach GPU accelerators
%(NVIDIA Tesla M2070)
(NVIDIA Tesla M2070 double precision peak 515 Gflops) 
available to this study.
}
The finite element discretization of the governing equations requires significant
expertise of the GPU computing architecture as well as detailed algorithmic
understanding of both the finite element technology and matrix-vector multiply
within the iterative linear system solver to maximize floating point operation
throughput and minimizes memory transactions latencies.
As opposed to explicitly storing and reading matrix entries
from global memory, the matrix-free method recalculates the local matrix entries
needed within the linear system solve.
This memory access design pattern has demonstrated a 4-10x speedup  over
the matrix explicit methods~\cite{Muller2013,Knepley_2_2013,Medina2014}.
Meanwhile, the steady state superposition analysis, Equation~\eqn{GreenSuperPosition},
provides treatment predictions with fewer floating point operations and would be selected under an Occam
razor~\cite{jaynes2003probability} philosophy highly preferential to simplicity.
While all kernels
for the present spectral element methods and preconditioned conjugate gradient
method were hand coded in this manuscript, library implementations of the
matrix-free iterative solver approach are also
appearing~\cite{Knepley2013,Knepley2014}.

@@

The finite element discretization of the governing equations, however, 
provides significant opportunity for further
physics-based improvements including 
higher order model spherical harmonic expansions in the laser fluence 
model~\cite{carp2004radiative}. 
As seen in Figure~\ref{QuantitativeComparison}, the {\color{red}chosen} models
introduce a bias in the model parameter
recovery that differs from published literature values.
{\color{red}The bias may arise from inaccuracies in the modeling assumptions, perhaps
most significant being the use of optical tissue properties that are
invariant in space and time. The literature has clear examples
where temperature/damage dependent and spatio-temporal dependent
parameters are critical to the
prediction~\cite{Mohammed05,Schutt2008}. The use of thermometry data during
only the test pulse is also expected to influence the recovered optical
properties.
The choice to use a single constant parameter for full time history of each
MRTI dataset is motivated by
the practical model training focus of this investigation.
Higher order physics
models of the fluence are be expected to provide the highest accuracy in
recovering the optical parameters during the calibration process and for
characterizing the tissue properties.
